---
title: "Sampling"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Terminology

An **element** is the entity on which data are collected.

A **population** is a collection of all the elements of interest.

A **sample** is a subset of the population.

The **sampled population** is the population from which the sample is
drawn

A **frame** is a list of elements from which the sample will be
selected.

A sample provides **estimates** of the population characteristics.

The question is whether the sample estimates are representative of the
population characteristics?

To answer this question, we need to ask what **sampling methods**
produce "reliable" estimates?

## Finite Population Sampling

Let $N$ be the population size where $N < \infty$ so $N$ is finite.

Let $n$ be the sample size where $n \le N$.

A simple random sample of size $n$ from a finite population of size $N$
is a sample selected such that each possible sample of size $n$ has the
same probability of being selected.

We can sample with replacement, however, sampling without replacement is
more typically used.

Remembering our discussion of combinations, we know that the number of
different samples of size $n$ that can be drawn from the population of
size $N$ is:

$$\frac{N!}{n!(N-n)!}$$

## Finite Sampling Example

Create a finite population of numbers from 0 to 100,000, sequenced by 1.

The population mean is

$$\mu = \sum_{i = 1}^N \frac{x_i}{N}$$

Generate 5 samples of size $n = 100$ without replacement.

The sample average is:

$$\bar{x} = \sum_{i = 1}^n \frac{x_i}{n}$$

The population mean is 50,000, however, the sample averages vary
"around" the population mean.

If you run the code multiple times, the population mean does not vary,
but the sample averages will change with each run as we have not
**set.seed** the code.

The code below does the following:

-   Creates a population of integers, sequenced by 1, from 0 to 100,000

-   Sets the samples variable to 5

-   Creates a vector, avg_x, filled with zeros with the number of
    columns set by samples

-   Runs a loop where the number of loops is set by the samples variable

-   In each loop, takes a random sample of 100 from the population and
    estimates the average

-   Creates a tibble of the averages and row binds the population mean

-   Transposes the vector of averages and the population mean

-   Produces a table of results

```{r, message = FALSE, warning = FALSE}

rm(list = ls())

library(dplyr)
library(kableExtra)

population <- seq(0, 100000, by = 1) 

samples <- 5
avg_x <- rep(0,samples)

for(i in 1:samples){
  avg_x[i] <- mean(sample(population, size = 100, replace = FALSE))
}

samples <- tibble(avg_x) %>% 
           rbind(mean(population)) %>% t()

kable(samples,
      align       = "c", 
      format.args = list(big.mark = ",", 
                         scientific = FALSE),
      caption     = "Sample Averages and Population Mean",
      col.names   = c("Sample 1", "Sample 2", 
                      "Sample 3", "Sample 4",
                      "Sample 5", "Population")) %>%
kable_styling(font_size = 10)

```

## Infinite Population Sampling

We can also sample from a population that continues to change over time.

$$N(t) < N(t+1) < N(t+2) < \dots$$

In this case, we cannot find a list of all the elements in the
population because the population is increasing without limit.

We cannot randomly select observations since the probability of
selection a sample of size $n$ is dependent upon when the population was
sampled.

If we consider population of students that have enrolled at ODU, this
number changes within each semester and across semesters. The population
of current and former students is not static and theoretically could
increase without bound.

To sample the ODU population we need to fulfill two conditions.

-   Each element in the random sample must come from the population of
    interest.

-   Each element in the random sample is selected independently.

## Infinite Sampling Example

Assume we want to construct a sample of 1,000 ODU students and estimate
average household income.

First, we need to ensure each element in the random sample comes from
the population of interest.

Our sampling method would need to randomly select ODU students from the
same time period.

In other words, the 2025 - 2026 Academic Year rather than across the
last two decades as our interest is estimating the household income of
the current student population.

Why?

If we sampled household income from 1970, would that be representative
of the household income of the current and future student population?

***In other words, currency of data matters.***

Second, we need to ensure each element is selected independently to
avoid selection bias.

We should not randomly sample from students at Webb Center or students
in the commuter car parking lot. Both sampling mechanicisms would
produce biased samples of the ODU population.

What if we found students who purchased from the bookstore during the
last week of October?

We could find each student in the registrar list and then select the
student $x$ number of positions alphabetically in front or behind the
student.

The purchases of students at the bookstore are random and independent of
other students. Thus, the selection of students $x$ positions in front
or behind in the alphabetical list of students yields a random sample
that is independently drawn from the infinite population.

## Estimators and Estimates

Point estimation is the use of sample data to estimate a sample
statistic.

**An estimator is a rule for calculating an estimate of a given quantity
based on observed data.**

-   The sample average, $\bar{x}$ is an estimator for the population
    mean.

-   The estimator $\bar{x}$ produces an estimate of the population mean
    $\mu$ using the $n$ observations,

-   $s^2$ is the estimator of the population parameter $\sigma^2$.

-   $s$ is the estimator of the population parameter $\sigma$.

-   $\bar{p}$ is the estimator of the population proportion $p$.

In the code below, we demonstrate how the sample average and sample
standard deviation are estimators of the population mean and the
population standard deviation.

We generate point estimates using:

$$\bar{x} = \sum_{i = 1}^n x_i / n$$

$$s = \sqrt{ \frac{\sum_{i = 1}^n(x_i - \bar{x})^2}{n-1}}$$

The code below does the following:

-   Generates 1,000,000 random numbers drawn from a normal distribution

-   The mean of the normal distribution is 25 with a standard deviation
    of 3

-   Declares 5 samples will be taken

-   Creates three vectors to be filled by the loop

-   A sample of 500 is taken from the population

-   The sample number is set to the iteration of the loop

-   The sample average and sample standard deviation are estimated

-   A tibble is created of all three variables

-   A table of output is produced

```{r, message = FALSE, warning = FALSE}

rm(list = ls())

library(dplyr)
library(kableExtra)

population <- rnorm(1000000, mean = 25, sd = 3) 

samples <-5 
samp_no <- rep(0,samples)
mean_vec <- rep(0,samples)
sd_vec <- rep(0,samples)

for(i in 1:samples){
  temp <- sample(population, size = 500, replace = FALSE)
  samp_no[i] <- i
  mean_vec[i] <- mean(temp)
  sd_vec[i] <- sd(temp)
}

samples <- tibble(samp_no, mean_vec,sd_vec)

kable(samples,
      align       = "ccc", 
      digits      = 2,
      caption     = "Sample and Population Averages",
      col.names   = c("Sample","Average", "SD")) %>%
kable_styling(font_size = 10)

```

An illustrative exercise is to increase the sample size relative to the
population and observe what happens to the estimates of the mean and
standard deviation.
