---
title: "Uniform Distributions"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    self-contained: false     # must be false when using webr
urlcolor: blue
filters:
  - webr
execute:
  webR: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo = FALSE, message = FALSE}

library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(tidyquant, quietly = TRUE)

```

## Discrete Random Variables

A **random variable** is a numerical description of an outcome of an
example.

A **discrete random variable** may assume a **finite** of **infinite**
number of values.

The **probability mass function** or **PMF**, $P(X)$, generates the
probability of $X$ occurring for a specific value of $x$.

The probability mass function is also referred to as the **discrete
probability density function**.

The **cumulative distribution function** or **CDF** generates the
probability of $X$ being less than or equal to a specific value of $x$.

The cumulative distribution function is also referred to as the
**discrete cumulative density function**.

Given a binomial experiment with the following properties, answer the
following questions using either the PDF or CDF.

-   $n = 50$

-   $p = .8$

-   $x = 42$

-   What is $P(X = x) = P(X = 42)$?

-   What is $P(X \le x) = P(X \le 42)$?

-   What is $P(X > x) = 1 - P(X \le X) = 1 - P(X \le 42)$?

```{r}

rm(list = ls())

library(dplyr)
library(kableExtra)

n = 50
p = .8
x = 42

probs <- tibble(p_1 <- dbinom(x, n, p),
                p_2 <- pbinom(x, n, p),
                p_3 <- pbinom(x, n, p, lower.tail = FALSE))

kable(probs,
      align = 'c',
      digits = 3,
      col.names = c('P(X = 42)', 'P(X <= 42)', 'P(X > 42)')) %>%
kable_classic()

```

## Continuous Random Variables

A **continuous random variable** may assume any value in an interval or
collection of intervals.

The **probability density function**, $f(x)$, is the counterpart of the
discrete probability density function.

The probability density function for a continuous random variable is
typically referred to as the **PDF**.

The PDF, $p(X)$, generates a curve for a given interval of $x$.

The PDF provides the probability that $x$ takes on any value in the
given interval of $x$.

This implies that the area under $p(X)$ for a specific value of $x$ is
zero.

In other words, given a continuous random value, the probability that
the continuous variable, $x$, is equal to a specific value is zero.

## Uniform Probability Distribution

The **continuous uniform probability distribution** describes an
experiment where there is a random outcome that is within specific
bounds.

The bounds of the random outcome are defined by the parameters, $a$ and
$b$.

-   $a$ is the minimum value of the random outcome.

-   $b$ is the maximum value of the random outcome.

The uniform distribution is defined by the parameters $a$ and $b$ which
define the region where the density function $p(x)$ is constant and
outside of which $p(x) = 0$. Since the area under the probability
density function must total 1, the probability density function of the
continuous uniform distribution is:

$$P(x | a,b)  = \frac{1}{b-a} \quad \text{for} \, x\in[a,b]$$

$$P(x |a,b) = 0 \quad \text{for} \, x \not\in[a,b]$$

We often denote the random variable $X$ as uniformly distributed on the
interval $[a,b]$ as:

$$
X \sim U(a,b)
$$

## Continuous Uniform Distribution

The PDF of the continuous uniform probability distribution is

$$f(x)  = \frac{1}{b-a} \quad \text{for} \, x\in[a,b]$$

$$\text{for} \, x \in[10,30]$$

For a continuous probability distribution, we know that:

$$f(X) \ge 0$$ $$\int_{-\infty}^{\infty} \, f(x) \, dx = 1$$

For the continuous uniform distribution:

$$a < b \rightarrow \frac{1}{b-a} \ge 0 \quad \forall \, a \le x \le b \rightarrow \, f(x) \ge 0 \quad \forall x$$
-

This means that $f(x) \ge 0$ for all $x \in [a,b]$.

Does the area defined by the PDF equal to one?

Here we can integrate the PDF and note that when $x < a$ and $x > b$
that $f(x) = 0$.

$$lim_{n \rightarrow \infty} \int_{-n}^{n} f(x)d(x) = lim_{n \rightarrow \infty} \int_{-n}^{a} (0)d(x) +\int_{a}^{b} \frac{1}{b-a}dx + \int_{b}^{n} (0)d(x)$$

$$lim_{n \rightarrow \infty} \int_{-n}^{n} f(x)d(x) = \frac{1}{b-a}[x]_{b}^{a} = \frac{1}{b-a}(b -a) = 1$$
-

The PDF of the continuous uniform probability distribution fulfills the
conditions for a PDF of a continuous random variable.
